---
title: "Programming with Functions"
author: "Jan Vitek"
date: "Wednesday May 3, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


- Basics: what is a function and why would you write one?
- How to write a function
- Parts of a function
    + Formals
    + Body
    + Environment
- Example: writing a function
    + Rewrite a previous example (from case study?) as a function
- Function arguments
    + Default and missing arguments
    + Lazy evaluation
- Returning values: implicit, `return()`, `invisible()`
- John Chambers: "To understand computations in R, two slogans are helpful:"
    + *Everything that exists is an object*
    + *Everything that happens is a function call*
- Example: `+` and `-`, etc., are also functions, can be redefined
- Infix functions, creating a user-defined `%plus%`, etc.
- How to write good functions



## Functions

The building block for creating abstractions in R is the *function declaration*.  These abstractions are used in *function calls*.

A function object is a value in the 

```{r, eval=FALSE}
f <- function( y ) {  y + 1 }
f( 1 )
```

Function definitions in R consist of an argument list and a body. The argument list is a sequence of variable names (with optional default values) and possibly one occurrence of the tripe dot symbol (`...`).  The body of the function can be a single expression or a block (`{...}`).

```{r, eval=FALSE}
# the shortest function in the book
function() 1    
# multiple expressions in the body
function( x ) { y <- x * x;  y + 1 }
# default values, optional arguments
function( x = 0, ..., y = 2)   if(length(z <- list(...) == 0)) x * y  else x * y * z[[1]] 
```

A function call consists of an expression that evaluates to a function followed by a (possibly empty) list of (possible named) arguments.

```{r, eval=FALSE}
f()
# the shortest call of them all
f(1,2)
# positional calls pass arguments in order
f(x = 1, y = 2)
f(y = 2, x = 1)
# named calls pass argument by name without a prescribed order
(x[42])(x=1)
# an expression that yields a function can be called
```

Some functions have inline syntax, `+` for example does not require parentheses and `if else` is  function which takes two arguments.  The assignment operators are functions `<-` and `[]<-, so are parenthese `(`.  All functions can be redefined.

```{r}
`+` <- function(a,b) paste(a,b)
`(` <- function(a) print("(" + a + ")")
(1 + 1)
```

**Exercise 1:** write a definition of a function `f` that will return `0`, `-1`, `1`, `1` for the four calls to `f` above and a variable `x` such that the last call returns `-1`.


Let's clean up the environment. 

```{r}
rm(`+`)
rm(`(`)
```


```{r, eval=F}
f <- function(y = 0, x = 0) x - y
f()
f(1,2)
f(x = 1, y = 2)
f(y = 2, x = 1)
```

The optional arguments have somewhat interesting semantics. In a function, any argument occurring after the three dots must be passed by name.  In the function body they can be passed as argument to a further call.

```{r, eval=F}
f <- function( x, ..., y) {  # y can only be passed by name
   l1 <- c(...)  # the tree dots become an argument list
   l2 <- c(x,...)
   l3 <- c(y,...,x)
   l1[1] + l2[2] + l3[3]
}
f(1,2,y=3)
```

**Exercise 2:** guess what value the call to `f` will return.

The rationale for having arguments after the three dots is to specify optional arguments rarely overridden.

Function evaluation is complete when the last expression in the body is evaluated. The `return()` call can break evaluation in the middle.  The `stop()` call can raise an exception that can only be caught in a `tryCatch()` statement. 

```{r, eval=F}
f <- function(x,acc) {
  if (x == 0) tryCatch( stop(simpleException("done")), 
                        simpleException = function(err) acc)
  else f(x-1, acc+1) 
}  
```


**Exercise 3:** what will happen if we evaluate `f(-1)`.

Each call to a function adds the argument to function on a data structure referred to as the *stack*. When a function returns the arguments are removed from the stack.


### Referential transparency

To facilitate equational reasoning, R attempts to provide *referential transparency* for function calls. Referential transparency means that arguments are not changed by the function being called. So the following function `f(x)` does not modify the vector passed in.

```{r}
x
f <- function(x) { x[1] <- 1 }
f(x)
x
```

But then what does it change in the body?

The `pryr` package provides function `object_size` to estimate the size of several data structures.

```{r}
X <- 1:(10*1000*1000)
if (!require("pryr",character.only = TRUE)) install.packages("pryr")
pryr::object_size(X)
Y <- X
pryr::object_size(X,Y)
Y[1] <- 11
pryr::object_size(X,Y)
```

**Exercise 9:** Explain the two surprising things about the above results.


## Delayed Evaluation

**Exercise 10:** Consider the following code, what will it print?

```{r}
print(1)   # outputs "1" on the terminal
f<-function(a,b) { invisible(0) }  # a helper, nevermind
f( print(1) , print(2) )   # what would this do in Java or in C
temp <-print(1);temp2 <- print(2); f(temp, temp2)  # same thing?
```

**Exercise 11:** What will this print?

```{r}
f1 <- function( a, b, c ) { # Consider the following, not very useful function
 x <- f(b,b)
 x <- c
 x <- c
 a
}

f1( print(1), print(2), print(3) ) # What do you recon this will print?
```

**Exercise 12:** What will this print?

```{r}
f2 <- function( a, b, c ) { # here is another function
  x <- b
  x <- c
  f3(a)
}

f3 <- function( a) { } # with a silly helper function

f2( print(1), print(2), print(3) ) # what will we print?
```

In R, arguments to a functions are not evaluated, they are packaged into *promises*. A promise is only evaluated if it is accessed (and only the first time it is accessed).  

Every expression is passed as a promise, except constants.

The following is an obscure corner of R.   Enjoy.

```{r}
FU <- function(b="") print(paste("FU",b))                # A function
BAR <- function( FU = print("BAR"))  FU("bar")   # Another function with a default argument
BAR()   # call bar will print ?
BAR( function(n) print("YO") )  # and now?
```


# Building Abstractions

We have now reviewed the basic building blocks of the language. Using these building blocks it is possible to construct powerful abstractions.

## Building a factor abstraction

First, lets look at how to implement factors using attributes.
 
```{r}
simple_factor <- function (x) {
    levels <- as.character(unique(x))
    f <- match(x, levels)
    attr(f, "levels") <- levels
    attr(f, "class") <- "simple_factor"
    f
}

x <- c(42, 42, 2, 3, 4)
simple_factor(x)
```
Printing is not very elegant. Let's add a dedicated print function.

```{r}
print.simple_factor <-  function (x, ...)  {
   xx <- x  # copy x
   class(xx) <- NULL  # delete its class
   levels(xx) <- NULL # delete its levels
   xx <- sapply(xx, function(y) attr(x,"levels")[y]) # replace levels with their values
   print(xx, quote=FALSE)   # print without quotes
   cat(paste("Levels: ", paste(levels(x),collapse=" ")))  #print the levels
  invisible(x)  # return x without echoing it
}

print(simple_factor(x))
```

## Building a counter abstraction

With referential transparency it is difficult to mutate non-local state.  What if we wanted to create a large number of counters and did not want anyone but us to be able to update them.



```{r}
remove(xPriv)
counter <- function() { xPriv <- xPriv + 1; xPriv}
#counter()             # oops, we need to define x
xPriv<-0
counter()
counter()             # Hmm... not quite

# Super assignment to the rescue
counter <- function() { xPriv <<- xPriv + 1; xPriv}  # <<- assigns in the parent env
counter()
counter()             # now we have a counter
```

This only lets us have a single counter. What if we want a counter factory?


```{r}
mkC <- function () { 
    xPriv <- 0
    function() { xPriv <<- xPriv + 1; xPriv }
  }

count <- mkC()
count()
count()
```

The above uses the environment of the `mkC()` function to store the counter. Each time we call `mkC`, a new environment is created.


## Build abstraction with domain specific languages

R allows to leverage delayed evaluation in order to implement domain specific languages.

Here is the example of timer function that repeatedly run the expression passed in as argument and reports the timing of that expression.

```{r}
tm <- function( exp ) {
 x<- deparse(substitute( exp ))   ## decompile the expression
 t <- 1                           ## repetion counter (we run exp 10 * t)
 time = 0;                        ## reset timer
 while ( time < 1 ) {             ## continue looking until the code runs long enough
  y <- x                          ## code
  for( i in 1:9 ) y <- paste(y,";",x)  ## duplicate the code 10 times
  y<- paste(" system.time ( for (i in 1:",t,") {", 
            y,                    ## wrap the code with a loop that runs t times
            "}, gcFirst= FALSE)")
  r <- eval(parse(text=y))        ## execute
  time <- r[3]                    ## grab the elasped time
  t <- t * 10                     ## try with a bigger t
 }
 (time / (t * 10)) * 1000         ## return time in milliseconds (default is in s)
}

tm(1 + 1)
```
